---
title: "EYSM example from the article"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
```

# Table of contents

1. [Introduction](#introduction)
2. [Short description](#paragraph1)
3. [Model description](#paragraph2)
4. [Plots](#plots)

## Introduction <a name="introduction"></a>

In this repo there is an example presented in the EYSM short paper "Resampling Methods in Conditional Independence Testing". The details are described in the article.


- `resampling.R` - implementation of the resampling methods
- `simulation_function.R`, `simulations.R` - simulation-related files
- `model.R` - implementation of the model described in the section [Model description](#paragraph2)
- `cpp_functions.cpp` - helper functions

## Short description <a name="paragraph1"></a>

Conditional mutual information is defined as
\[CMI(p) = I(X,Y|Z)=\sum_{x,y,z}p(x,y,z)\log\frac{p(x,y,z)p(z)}{p(x,z)p(y,z)}.\]

The estimator of a vector of probabilities $(p(x,y,z))_{x,y,z}$ is a vector of fractions $(\hat p(x,y,z))_{x,y,z} = (n(x,y,z)/n)_{x,y,z}$, where $n(x,y,z) = \sum_{i=1}^n\mathbb{I}(X_i=x, Y_i=y, Z_i=z)$. 
We estimate conditional mutual information
 using a plug-in estimator, namely 
\[CMI(\hat p) = \sum_{x,y,z} \hat p(x,y,z) \log \frac{\hat p(x,y,z) \hat p(z)}{\hat p(x,z) \hat p(y,z)}.\]

**Lemma 1**

If $X \perp \!\!\! \perp Y | Z$ we have that $2n CMI(\hat p) \xrightarrow{d} \chi_{(|\mathcal{X}|-1)(|\mathcal{Y}|-1)|\mathcal{Z}|}^2$.


**Theorem 2**

If the null hypothesis $H_0: X \perp \!\!\! \perp Y |Z$ holds, then
\[P\left(\frac{1 + \sum_{b=1}^B \mathbb{I}(T \leq T_b^*)}{1+B} \leq \alpha \right) \leq \alpha,\]
where $T=T(\mathbf{X}_n, \mathbf{Y}_n, \mathbf{Z}_n)$ and $T_b^* = T(\mathbf{X}_{n,b}^*, \mathbf{Y}_{n,b}^*, \mathbf{Z}_{n,b}^*)$,
$(\mathbf{X}_n, \mathbf{Y}_n, \mathbf{Z}_n)$ is a sample and $(\mathbf{X}_{n,b}^*, \mathbf{Y}_{n,b}^*, \mathbf{Z}_{n,b}^*)$ is a resampled sample (CR or CP scenarios).

### Tests

We consider three tests for testing conditional independence $H_0: X \perp \!\!\! \perp Y|Z$: 

- `asymptotic` - a test based on Lemma 1,
- `exact` - a test based on Theorem 2,
- `df estimation` - a test that uses the chi-squared distribution as a benchmark distribution, but adjusts the number of degrees of freedom $d$ based on resampled samples, $\hat d = \frac{1}{B}\sum_b \widehat{CMI}_b^*$.

## Model description <a name="paragraph2"></a>

Joint probability in the model is given as
    \[p(x,y,z_1, z_2, z_3, z_4) = p(y)p(x|y) \prod_{s=1}^4p(z_i|y),\]
    and is presented in Figure 1.
   $Y$ is a Bernoulli random variable with probability of success equal to $0.5$ and conditional distribution of 
    $\tilde{X}$ and $\tilde{Z}_i$ for $i=1,2,3,4$ given $Y=y$  follows a normal distribution: $\tilde{X}|Y=y \sim N(y, 1)$, $\tilde{Z_i}|Y=y \sim N(\gamma^iy, 1)$ and $\gamma\in [0, 1]$ is a parameter. In this example, $\gamma=0.5$. In order to obtain discrete variables from continuous $(\tilde{X}, \tilde{Z}_1, \tilde{Z}_2, \tilde{Z}_3,\tilde{Z}_4)$ we define 
    \[P(X=x\lvert Y=y) = P\Big((-1)^{x}\tilde{X}\leq\frac{(-1)^{x}}{2}\lvert Y=y\Big),\]
    \[P(Z_i=z_i\lvert Y=y) = P\Big((-1)^{z_i}\tilde{Z}_i\leq\frac{(-1)^{z_i}\gamma^i}{2}\lvert Y=y\Big)\]
    for $i = 1, 2, 3,, 4$, where $x, z_1, z_2, z_3, z_4 \in \{0,1\}$. Thus $X\lvert Y=y \sim Bern(\Phi((2y-1)/2)$ and $Z_i\lvert Y=y \sim Bern(\Phi(\gamma^i(2y-1)/2))$.  Variables $X, Z_1, Z_2, Z_3, Z_4$ are conditionally independent given $Y$ but $X$ an $Y$ are not conditionally independent given $Z_1,Z_2,Z_3, Z_4$.

## Plots <a name="plots"></a>

```{r}
results <- read.csv("significance.csv")

library(dplyr)
library(ggplot2)

results[,"rejected"] <- (results[,"pvalue"] < 0.05)
df <- results %>% 
  group_by(resampling, test, n, s, frac) %>%
  summarise(power=mean(rejected))

df_resampling_tmp <- df %>% filter(resampling %in% c("BCI", "BX"), test != "asymptotic") %>%
  mutate(name=paste0(test, " (", resampling, ")"))
df_asymptotic_tmp <- df %>% filter(resampling %in% c("BCI"), test == "asymptotic") %>%
  mutate(name=paste0(test))

df <- rbind(df_resampling_tmp, df_asymptotic_tmp)

df$name <- factor(df$name, levels=c("asymptotic", "exact (BCI)", "df_estimation (BCI)", "exact (BX)", "df_estimation (BX)"))

plot_colors <- RColorBrewer::brewer.pal(10, "Paired")
plot_colors_1 <- plot_colors[c(10,3:4,7:8)]
plot_colors_2 <- plot_colors[c(1:2, 5:6)]

p <- ggplot(df, aes(x=frac, y=power, shape=name, color=name)) +
  geom_line() +
  geom_point() +
  scale_color_manual(values=c(plot_colors_1), name="Test") +
  scale_shape_manual(values=1:5, name="Test") +
  geom_hline(yintercept=0.05, linetype="dotted") +
  geom_hline(yintercept=0.1, linetype="dashed") +
  theme_bw() +
  ylab("Fraction of rejections") +
  ggtitle("Significance level", subtitle="Tests exceeding the significance level") +
  theme(legend.position = c(1-0.02,1-0.02), legend.justification = c(1,1))
ggsave("significance_v1.pdf", p, width=4, height=4)
ggsave("significance_v1.png", p, width=4, height=4)
p
```
```{r}
results <- read.csv("significance.csv")

library(dplyr)
library(ggplot2)

results[,"rejected"] <- (results[,"pvalue"] < 0.05)
df <- results %>% 
  group_by(resampling, test, n, s, frac) %>%
  summarise(power=mean(rejected))

df_resampling_tmp <- df %>% filter(resampling %in% c("CR", "CP"), test != "asymptotic") %>%
  mutate(name=paste0(test, " (", resampling, ")"))

df <- df_resampling_tmp

df$name <- factor(df$name, levels=c("exact (CR)", "df_estimation (CR)", "exact (CP)", "df_estimation (CP)"))

plot_colors <- RColorBrewer::brewer.pal(10, "Paired")
plot_colors_1 <- plot_colors[c(10,3:4,7:8)]
plot_colors_2 <- plot_colors[c(1:2, 5:6)]

p <- ggplot(df, aes(x=frac, y=power, shape=name, color=name)) +
  geom_line() +
  geom_point() +
  scale_color_manual(values=c(plot_colors_2), name="Test") +
  scale_shape_manual(values=1:5, name="Test") +
  geom_hline(yintercept=0.05, linetype="dotted") +
  geom_hline(yintercept=0.1, linetype="dashed") +
  theme_bw() +
  ylab("Fraction of rejections") +
  ggtitle("Significance level", subtitle="Tests holding the significance level") +
  ylim(0, 0.2) +
  theme(legend.position = c(1-0.02,1-0.02), legend.justification = c(1,1))
ggsave("significance_v2.pdf", p, width=4, height=4)
ggsave("significance_v2.png", p, width=4, height=4)
p
```

```{r}
results <- read.csv("power.csv")

library(dplyr)
library(ggplot2)

results[,"rejected"] <- (results[,"pvalue"] < 0.05)
df <- results %>% 
  group_by(resampling, test, n, s, frac) %>%
  summarise(power=mean(rejected))

df_resampling_tmp <- df %>% filter(resampling %in% c("CR", "CP"), test != "asymptotic") %>%
  mutate(name=paste0(test, " (", resampling, ")"))

df <- df_resampling_tmp

df$name <- factor(df$name, levels=c("exact (CR)", "df_estimation (CR)", "exact (CP)", "df_estimation (CP)"))

plot_colors <- RColorBrewer::brewer.pal(10, "Paired")
plot_colors_1 <- plot_colors[c(10,3:4,7:8)]
plot_colors_2 <- plot_colors[c(1:2, 5:6)]

p <- ggplot(df, aes(x=frac, y=power, shape=name, color=name)) +
  geom_line() +
  geom_point() +
  scale_color_manual(values=c(plot_colors_2), name="Test") +
  scale_shape_manual(values=1:5, name="Test") +
  theme_bw() +
  ylab("Fraction of rejections") +
  ggtitle("Power", subtitle="Tests holding the significance level") +
  theme(legend.position = c(1-0.02,0+0.02), legend.justification = c(1,0))
ggsave("power.pdf", p, width=4, height=4)
ggsave("power.png", p, width=4, height=4)
p
```